---
title: Age-related gene expression variations <br> in human monocytes and T cells
author:
- Kruthika Iyer
- Cristian Valencia
- Martin Skarzynski
date: "`r Sys.Date()`"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE,
                      include = TRUE, warning = FALSE, message = FALSE)
```

```{r libs}
library(readr)
library(here)
library(plotly)
library(knitr)
```

# Context

# Packages

Processing

- `beadarray` R package
    - QC analyses and bead-type summarization (average bead signal for each type after outlier removal)
- `limma` R package - normal-exponential convolution model analysis to estimate non-negative signal,
    - quantile normalization using all probes (gene and control, detected and not detected) and samples,
    - addition of a recommended (small) offset,
    - log2 transformation and elimination of control probe data from the normalized expression matrix.
- `lumi` package
    - Illumina HumanMethylation450 BeadChip technology employs a two-channel system and uses both Infinium I and II assays; normalization was performed.

Association

- `Stats` package
- `MASS` package

# Goals

# Data Processing

# Results Summary

# Conclusions
